---
title: "An introduction to the caliver package"
author: "Claudia Vitolo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: references.bib
vignette: >
  %\VignetteIndexEntry{caliver}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      eval = FALSE)
```

The name caliver stands for **CALI**bration and **VER**ification of gridded model outputs. It is a package developed for the R programming language and available with an APACHE-2 license from a public repository. The caliver package was initially designed for the post-processing, calibration and validation/verification of Forest Fire models but as development progressed, it became clear that it could have a much wider applicability. The goal of this vignette is to describe the functionalities of the package, examples are given with regard to outputs of the following forest fire models: GEFF and RISICO. Applicability to other model results will be investigated in future works. Complete documentation is available within the package. This vignette contains the code to run all the examples in [@vitolo:2017b], please refer to this paper for more information.

## Package installation and dependencies

The caliver package is implemented in the R statistical language and depends on the following external libraries: the Climate Data Operators [@schulzweida2006cdo] a large tool set for working on climate and NWP model data), the NCAR Command Language [@brown2012ncar] an interpreted language for scientific data analysis, the Geospatial Data Abstraction Library [@warmerdam2008geospatial] a translator library for raster and vector geospatial data formats, and the NetCDF4 library [@rew1990netcdf].

The caliver package also depends on additional R packages from both CRAN and Bioconductor. Users must have the above dependencies installed before attempting to install caliver. Installation instructions for different operating systems are available in the README file.

```{r, echo = FALSE, eval = FALSE}
setwd('/scratch/mo/moc0/fire/')
```

Once all the dependencies are available, the package can be installed via [devtools](https://github.com/hadley/devtools):
``` r
devtools::install_github("ecmwf/caliver")
```

Load the caliver package:

``` r
library("caliver")
```

## Get data from the fourth-generation Global Fire Emissions Database (GFED4)

Observed burned areas around the world are collected by the GFED4 in hdf format \citep{giglio2013analysis}. This information is very important as it represents the ground truth for fire models, it is needed to make comparisons with reanalysis and forecast models and estimate their reliability.

``` r
# Get daily burned area maps from 2003 to 2015 (to be run in the console!)
BurnedAreas <- get_gfed4(start_date = "2003-01-01",
                         end_date = "2015-12-31",
                         temporal_resolution = "daily",
                         varname = "BurnedArea",
                         region = "GLOB")
```

```{r, echo = FALSE, eval = FALSE}
# The above can be saved and re-loaded as follows:
raster::writeRaster(BurnedAreas,
                    filename="BurnedArea.grd",
                    bandorder='BIL', overwrite=TRUE, progress = 'text')
BurnedAreas <- raster::brick("BurnedArea.grd")
```

The fourth-generation global fire emissions database also provides as ancillary data a map of 14 basis regions used to evaluate regional annual emission estimates. This map can be retrieved as SpatialPolygonsDataFrame (a spatial class defined in the sp package) setting `varname = "BasisRegions"'.

``` r
# Get all the BasisRegions
BasisRegions <- get_gfed4(varname = "BasisRegions")

# Europe
Europe <- get_gfed4(varname = 'BasisRegions', region = 'EURO')
```

Get administrative areas using getdata from the raster package
``` r
# United Kingdom
UnitedK <- raster::getData(name = "GADM", country = "United Kingdom", level = 0)

# Portugal
# PT <- raster::getData(name = "GADM", country = "Portugal", level = 0)

# Spain
Spain <- raster::getData(name = "GADM", country = "Spain", level = 0)

# Italy
Italy <- raster::getData(name = "GADM", country = "Italy", level = 0)

# Liguria Region
Italy1 <- raster::getData(name = "GADM", country = "Italy", level = 1)
Liguria <- Italy1[9,]

# Calabria Region
Calabria <- Italy1[4,]

# Sicily
Sicily <- Italy1[15,]

# Province of Genoa
Italy2 <- raster::getData(name = "GADM", country = "Italy", level = 2)
Genoa <- Italy2[42,]
rm(Italy1, Italy2)
```

## Utilities

Decompress archives using the function `decompress_gz()`.

``` r
decompress_gz(input_dir = "./tmp")
```

Once all the files are decompressed they can be merged in a unique netCDF file using the function `stack_netcdf_files()`.
```r
processingTime <- system.time({
  stack_netcdf_files(input_dir = 'GEFF1.2/fwi_rotated/',
                     output_file = 'FWIr_1980-2016.nc')
})
```

A simple benchmark
``` r
library(microbenchmark)
raster::writeRaster(s, filename='inst/testdata/GEFF.nc', format="CDF", overwrite=TRUE)

microbenchmark(
  sqrt(x),
  x ^ 0.5
)

```

## Analysis and visualisation

A first understanding of how FWI value are distributed over time can be achieved by creating a map of percentiles. In the example below we generate two maps: 1 for the median and the othe for the 90th percentile. 

``` r
maps <- makePercentileRaster(inFilePath = "FWIr_1980-2016.nc",
                             probs = c(50, 90))
```

### Mask/crop/subset

FWI values are related to the local vegetation as well as to localised soil and atmospheric conditions during the dry season. For this reason it is often useful to look at spatio-temporal subsets of the full reanalysis dataset. In the example below, the function `mask_crop_subset()` is used to extract the FWI for the first data of the reanalysis (setting argument `idx = 1`), masked and cropped over Europe.

``` r
mapEurope <- mask_crop_subset(r = maps, p = Europe, mask = T, crop = T, idx = 1)
```

### Plot

The function `plotPercentiles()` provides a modified version of the `raster::plot()` function. Below is a comparison.

``` r
# Use the raster plot method
setEPS()
postscript("~/raster.eps", width = 5.2, height = 2.7)
raster::plot(maps, main = c("FWI 50th perc.", "FWI 90th perc."))
dev.off()

# Use the caliver plotPercentiles function
setEPS()
postscript("~/caliver.eps", width = 5.2, height = 3)
plotPercentiles(maps, main = c("FWI 50th perc.", "FWI 90th perc."))
dev.off()
```

### Assemble the FWI dataset

In order to calculate the danger classes, the FWI brick should be loaded and the indices corresponding to the local fire season should be identified. In the example below the fire season is assumed constant across Europe, starting in April and ending in October each year.

``` r
# Define period for Reanalysis
dataDates <- seq.Date(from = as.Date('1980-01-01'),
                      to = as.Date('2016-12-31'),
                      by = "day")

# Define a function to extract fire seasons in Europe
seasons <- get_fire_season(DATES = dataDates, zone = 'north')

# Create an index of summer dates
fireSeasonIndex <- which(seasons == TRUE)

# Load, mask and subset the FWI dataset obtained previously
FWI <- raster::brick('FWIr_1980-2016.nc')
```

### Calculate danger levels

Threshold for every area of interest are calculate on the subset of the FWI brick related to the fire season only and cropped over the area of interest. Below are examples calculated for various countries in Europe, regions and a province in Italy. The result shows that fire danger classes is generally correlated to the local climate and spatial variability appears to be meaningful up to regional level. 

``` r
# Mask/Crop/Subset FWI and generate thresholds for Europe
EURO <- mask_crop_subset(r = FWI, p = Europe, mask = T, crop = T, 
                    idx = fireSeasonIndex)
EuropeThr <- FireDangerLevels(fire_index = EURO)

# Mask/Crop/Subset FWI and generate thresholds for United Kingdom
UK <- mask_crop_subset(r = FWI, p = UnitedK, mask = T, crop = T, 
                  idx = fireSeasonIndex)
UnitedKThr <- FireDangerLevels(fire_index = UK)

# Mask/Crop/Subset FWI and generate thresholds for Spain
ES <- mask_crop_subset(r = FWI, p = Spain, mask = T, crop = T, 
                  idx = fireSeasonIndex)
SpainThr <- FireDangerLevels(fire_index = ES)

# Mask/Crop/Subset FWI and generate thresholds for Italy
IT <- mask_crop_subset(r = FWI, p = Italy, mask = T, crop = T, 
                    idx = fireSeasonIndex)
ItalyThr <- FireDangerLevels(fire_index = IT)

# Mask/Crop/Subset FWI and generate thresholds for Calabria
CAL <- mask_crop_subset(r = FWI, p = Calabria, mask = T, crop = T, 
                  idx = fireSeasonIndex)
CALThr <- FireDangerLevels(fire_index = CAL)

# Mask/Crop/Subset FWI and generate thresholds for Sicily
SIC <- mask_crop_subset(r = FWI, p = Sicily, mask = T, crop = T, 
                  idx = fireSeasonIndex)
SICThr <- FireDangerLevels(fire_index = SIC)

# Mask/Crop/Subset FWI and generate thresholds for Liguria
LIG <- mask_crop_subset(r = FWI, p = Liguria, mask = T, crop = T, 
                  idx = fireSeasonIndex)
LIGThr <- FireDangerLevels(fire_index = LIG)

# Mask/Crop/Subset FWI and generate thresholds for Genoa
GEN <- mask_crop_subset(r = FWI, p = Genoa, mask = T, crop = T, 
                   idx = fireSeasonIndex)
GENThr <- FireDangerLevels(fire_index = GEN)
```

### Plot density with thresholds

The thresholds are different fromt the percentiles, the PDF  below shows a comparison for Italy.

``` r
countryPDF <- plotPDF(fire_index = IT, countryName = "Italy", 
                      ItalyThr, upperLimit = 75, vLines = c(0.50, 0.75, 0.90))
```

### Validate danger levels

In order to test whether the classes identified correspond to 

``` r
BurnedAreas <- raster::brick("GFED4_BurnedAreas/BurnedArea.grd")

# Mask and crop burned areas over Europe
BA <- mask_crop_subset(r = BurnedAreas, p = Europe, mask = T, crop = T)

# If observations layers have no date, assign it!
names(BA) <- seq.Date(from = as.Date("2003-01-01"),
                      to = as.Date("2015-12-31"), by = "day")

# For the validation we do not want to subset over the fire season, subset to match days in BA
EURO2 <- mask_crop_subset(r = FWI, p = Europe, mask = T, crop = T, 
                    idx = which(names(FWI) %in% names(BA)))

# Generate the contingency table for Europe
x <- ValidateFireDangerLevels(fire_index = EURO2, 
                              observation = BA,
                              fireThr = 10, obsThr = 50)
                         
# Contingency table
y <- data.frame(round(prop.table(x),3))


hits <- x$Freq[which(x$FWIlogic == TRUE & x$BAlogic == TRUE)]
misses <- x$Freq[which(x$FWIlogic == FALSE & x$BAlogic == TRUE)]
falsealarms <- x$Freq[which(x$FWIlogic == TRUE & x$BAlogic == FALSE)]
correctneg <- x$Freq[which(x$FWIlogic == FALSE & x$BAlogic == FALSE)]

POD <- hits/(hits+misses)               # 66%
FAR <- falsealarms/(hits+falsealarms)   # 98%
FBI <- (hits+falsealarms)/(hits+misses) # 38%
TS <- hits/(hits+misses+falsealarms)    # 1.7%

# Generate the contingency table for Europe using JRC's high level threshold
xJRC <- ValidateFireDangerLevels(fire_index = EURO2, 
                                 observation = BA,
                                 fireThr = 21.3, obsThr = 50)
# Contingency table
yJRC <- data.frame(round(prop.table(xJRC),3))

# Repeat the above for UK
BA_UK <- mask_crop_subset(r = BurnedAreas, p = UnitedK, mask = T, crop = T)
names(BA_UK) <- seq.Date(from = as.Date("2003-01-01"),
                         to = as.Date("2015-12-31"), by = "day")
UK2 <- mask_crop_subset(r = FWI, p = UnitedK, mask = T, crop = T, 
                   idx = which(names(FWI) %in% names(BA_UK)))
xUK_caliver <- data.frame(ValidateFireDangerLevels(fire_index = UK2, 
                                                   observation = BA_UK,
                                                   fireThr = 10, obsThr = 50))
xUK_JRC <- data.frame(ValidateFireDangerLevels(fire_index = UK2, 
                                               observation = BA_UK,
                                               fireThr = 21.3, obsThr = 50))
                                                   
# Repeat the above for Spain
BA_ES <- mask_crop_subset(r = BurnedAreas, p = Spain, mask = T, crop = T)
names(BA_ES) <- seq.Date(from = as.Date("2003-01-01"),
                         to = as.Date("2015-12-31"), by = "day")
ES2 <- mask_crop_subset(r = FWI, p = Spain, mask = T, crop = T, 
                   idx = which(names(FWI) %in% names(BA_ES)))
xES_caliver <- data.frame(ValidateFireDangerLevels(fire_index = ES2, 
                                                   observation = BA_ES,
                                                   fireThr = 10, obsThr = 50))
xES_JRC <- data.frame(ValidateFireDangerLevels(fire_index = ES2, 
                                               observation = BA_ES,
                                               fireThr = 21.3, obsThr = 50))
                                                   
# Repeat the above for Italy
BA_IT <- mask_crop_subset(r = BurnedAreas, p = Italy, mask = T, crop = T)
names(BA_IT) <- seq.Date(from = as.Date("2003-01-01"),
                         to = as.Date("2015-12-31"), by = "day")
IT2 <- mask_crop_subset(r = FWI, p = Italy, mask = T, crop = T, 
                   idx = which(names(FWI) %in% names(BA_IT)))
xIT_caliver <- data.frame(ValidateFireDangerLevels(fire_index = IT2, 
                                                   observation = BA_IT,
                                                   fireThr = 10, obsThr = 50))
xIT_JRC <- data.frame(ValidateFireDangerLevels(fire_index = IT2, 
                                               observation = BA_IT,
                                               fireThr = 21.3, obsThr = 50))
```
